---
title: "Customer Churn Prediction — Logistic Regression"
subtitle: "A Teaching Notebook | R Studio"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,    # show all code
  message = FALSE,   # suppress package messages
  warning = FALSE,   # suppress warnings for cleaner output
  fig.width  = 9,
  fig.height = 5
)
```

---

# 🎯 Objective

We want to predict whether a telecom customer will **churn** (cancel their service) using historical account and usage data.

This is a **binary classification** problem:

- `1` → Customer **churned**
- `0` → Customer **did not churn**

Because the outcome is binary/categorical, we use **Logistic Regression** — not linear regression.

> **Why not linear regression?**  
> Linear regression can predict values outside [0, 1], making it inappropriate for probabilities.  
> Logistic regression wraps the linear predictor in a **sigmoid (logistic) function**:  
> `P(churn=1) = 1 / (1 + e^(-z))` — output always between 0 and 1. ✓

---

## 🗺️ Roadmap

1. Load libraries and data  
2. Exploratory Data Analysis (EDA)  
3. Ensure the response variable is categorical  
4. Feature selection (correlation-based)  
5. Normalise features (critical for convergence)  
6. Train the logistic regression model  
7. Evaluate using the loss function (log-loss) + metrics  
8. Interpret the results  

---

# Step 1 — Load Libraries

We load everything upfront. In R, `tidyverse` handles data wrangling and plotting, `caret` handles the machine learning pipeline.

```{r load-libraries}
# ── Core data wrangling & visualisation ─────────────────────────────────────
library(tidyverse)     # dplyr, ggplot2, readr, tidyr, etc.
library(corrplot)      # correlation matrix visualisation
library(scales)        # nice axis labels

# ── Machine learning pipeline ────────────────────────────────────────────────
library(caret)         # train/test split, preprocessing, model training

# ── Evaluation ───────────────────────────────────────────────────────────────
library(pROC)          # ROC curves and AUC
library(MLmetrics)     # LogLoss function

# ── Table formatting ─────────────────────────────────────────────────────────
library(knitr)         # kable() for pretty tables
library(kableExtra)    # enhanced kable styling

cat("✅ All libraries loaded successfully.\n")
```

---

# Step 2 — Load & Explore the Data

Before building any model, we **understand our data**. This dataset contains **200 telecom customers** and **28 columns** of usage, demographic, and service information.

```{r load-data}
# Load the CSV — ensure ChurnData.csv is in your working directory
# Use getwd() to check your working directory, setwd() to change it
df <- read_csv("ChurnData.csv")

cat("Dataset dimensions:", nrow(df), "rows ×", ncol(df), "columns\n\n")

# Preview the first few rows
head(df) %>%
  kable(caption = "First 6 rows of ChurnData") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r eda-structure}
# Check the structure — equivalent to Python's df.info()
# Notice all columns are <dbl> (numeric floats)
glimpse(df)
```

```{r eda-summary}
# Summary statistics — min, max, mean, median, quartiles
# Notice wide ranges: income goes up to ~1600, whereas 'wireless' is binary
# This is precisely WHY we normalise — to put everything on the same scale
summary(df)
```

```{r eda-missing}
# Check for missing values
missing_counts <- colSums(is.na(df))
if (sum(missing_counts) == 0) {
  cat("✅ No missing values found in any column.\n")
} else {
  cat("Missing values per column:\n")
  print(missing_counts[missing_counts > 0])
}
```

---

# Step 3 — Response Variable: Make it Categorical

**Why does this matter?**

Logistic regression in R's `glm()` treats the response as a **factor** (R's categorical type).  
If we leave `churn` as a numeric float (e.g. `0.0` / `1.0`):

- `glm()` will try to run a *linear* regression, not logistic
- You'd receive a warning: *"In Ops.factor ... non-numeric argument"*
- The model would be fundamentally wrong

We must convert `churn` to a **factor with levels 0 and 1**.

```{r response-variable}
# ── Check current type ───────────────────────────────────────────────────────
cat("Current type of 'churn':", class(df$churn), "\n")
cat("Unique values            :", unique(df$churn), "\n\n")

# ── Cast to factor (categorical) ─────────────────────────────────────────────
# 0 = did not churn, 1 = churned
# Reference level (0) is set automatically as the first level
df$churn <- factor(df$churn, levels = c(0, 1), labels = c("No", "Yes"))

cat("After casting:\n")
cat("New type     :", class(df$churn), "\n")
cat("Levels       :", levels(df$churn), "\n\n")
print(table(df$churn))
```

```{r class-balance-plot}
# Visualise class balance — is the dataset imbalanced?
# Imbalanced classes (e.g. 95% No, 5% Yes) can bias the model toward the majority class

churn_counts <- as.data.frame(table(df$churn))
names(churn_counts) <- c("Churn", "Count")
churn_counts$Pct <- churn_counts$Count / sum(churn_counts$Count)
churn_counts$Label <- c("Did Not Churn (No)", "Churned (Yes)")

ggplot(churn_counts, aes(x = Label, y = Count, fill = Label)) +
  geom_col(width = 0.5, colour = "white") +
  geom_text(aes(label = paste0(Count, "\n(", percent(Pct, accuracy = 0.1), ")")),
            vjust = -0.4, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title    = "Churn Class Distribution",
    subtitle = "Target variable balance check",
    x = NULL, y = "Number of customers"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# Step 4 — Feature Selection Using Correlation

**Why select features carefully?**

Including irrelevant features:

- Adds noise → hurts prediction accuracy  
- Inflates the model → harder to interpret  
- Can cause **multicollinearity** → correlated predictors distort coefficients  

**Our strategy:**  
1. Compute point-biserial correlation (equivalent to Pearson for binary targets) of each numeric feature with the numeric `churn` values  
2. Keep features where |correlation| > 0.10  
3. Inspect a heatmap for multicollinearity among the selected features  

```{r feature-correlation}
# Convert churn back to numeric temporarily for correlation computation
churn_numeric <- as.numeric(df$churn) - 1   # Yes=1, No=0

# Compute correlations of all numeric features with churn
numeric_cols <- df %>%
  select(where(is.numeric)) %>%
  names()

corr_with_churn <- sapply(numeric_cols, function(col) {
  cor(df[[col]], churn_numeric, use = "complete.obs")
})

# Sort by absolute value
corr_df <- tibble(
  Feature     = names(corr_with_churn),
  Correlation = corr_with_churn
) %>%
  arrange(desc(abs(Correlation)))

cat("Pearson Correlation with 'churn' (sorted by |r|):\n")
corr_df %>%
  mutate(Correlation = round(Correlation, 3)) %>%
  kable(caption = "Feature correlations with churn") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r correlation-barplot}
# Visualise as a horizontal bar chart
corr_df_plot <- corr_df %>%
  mutate(Direction = ifelse(Correlation > 0, "Positive (+)", "Negative (-)"))

ggplot(corr_df_plot, aes(x = Correlation, y = reorder(Feature, abs(Correlation)),
                          fill = Direction)) +
  geom_col(colour = "white") +
  geom_vline(xintercept =  0.1, linetype = "dashed", colour = "forestgreen", linewidth = 0.8) +
  geom_vline(xintercept = -0.1, linetype = "dashed", colour = "forestgreen", linewidth = 0.8) +
  scale_fill_manual(values = c("Negative (-)" = "steelblue", "Positive (+)" = "tomato")) +
  labs(
    title    = "Feature Correlation with Churn",
    subtitle = "Dashed lines = ±0.10 selection threshold",
    x = "Pearson Correlation Coefficient",
    y = NULL,
    fill = "Direction"
  ) +
  theme_minimal(base_size = 12)
```

```{r select-features}
# ── Apply threshold ──────────────────────────────────────────────────────────
THRESHOLD <- 0.10

selected_features <- corr_df %>%
  filter(abs(Correlation) > THRESHOLD) %>%
  pull(Feature)

cat(sprintf("Features selected (|r| > %.2f): %d\n", THRESHOLD, length(selected_features)))
cat(paste(selected_features, collapse = ", "), "\n")
```

```{r multicollinearity-heatmap}
# ── Check multicollinearity among selected features ──────────────────────────
# If two predictors are highly correlated WITH EACH OTHER, including both
# is redundant and distorts coefficient estimates.

feat_corr_matrix <- df %>%
  select(all_of(selected_features)) %>%
  cor(use = "complete.obs")

corrplot(
  feat_corr_matrix,
  method   = "color",
  type     = "lower",
  tl.cex   = 0.75,
  addCoef.col = "black",
  number.cex  = 0.55,
  col      = colorRampPalette(c("steelblue", "white", "tomato"))(200),
  title    = "Feature–Feature Correlation (Multicollinearity Check)",
  mar      = c(0, 0, 2, 0)
)
```

```{r remove-collinear}
# ── Remove log-transformed duplicates ────────────────────────────────────────
# 'loglong' is a log-transform of 'longmon' — they carry the same information.
# 'lninc' is a log-transform of 'income' — same.
# Keeping both inflates variance of coefficients (multicollinearity).
# We keep the more interpretable original variables.

features_to_drop  <- c("loglong", "lninc")
selected_features <- setdiff(selected_features, features_to_drop)

cat(sprintf("Final feature set (%d features):\n", length(selected_features)))
corr_df %>%
  filter(Feature %in% selected_features) %>%
  arrange(desc(abs(Correlation))) %>%
  mutate(Correlation = round(Correlation, 3),
         Direction = ifelse(Correlation > 0, "▲ increases churn", "▼ decreases churn")) %>%
  kable(caption = "Final selected features with churn correlation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Step 5 — Normalisation (Adjusting for Convergence)

**Why normalise?**

Logistic regression in R's `glm()` uses **Iteratively Reweighted Least Squares (IRLS)**, which is sensitive to feature scale.

- `income` can exceed 1000  
- `wireless` is binary (0 or 1)

Without normalisation, the optimiser struggles with gradient steps of wildly different magnitudes — leading to **slow or failed convergence**.

**`preProcess(method = "scale", "center")`** from `caret` applies StandardScaling:  
`z = (x - μ) / σ` → Mean = 0, SD = 1

> ⚠️ **Critical rule:** Fit the scaler on **training data only**, then apply to both train and test. Fitting on test data leaks information ("data leakage").

```{r train-test-split}
# ── Prepare the modelling data frame ─────────────────────────────────────────
model_df <- df %>%
  select(all_of(c(selected_features, "churn")))

# ── Stratified train/test split (80/20) ──────────────────────────────────────
# Stratified = preserves the churn proportion in both splits
set.seed(42)  # for reproducibility

train_idx <- createDataPartition(
  model_df$churn,
  p     = 0.80,
  list  = FALSE,
  times = 1
)

train_df <- model_df[ train_idx, ]
test_df  <- model_df[-train_idx, ]

cat("Training set :", nrow(train_df), "samples\n")
cat("Test set     :", nrow(test_df),  "samples\n\n")
cat("Churn rate in train:", percent(mean(train_df$churn == "Yes"), 0.1), "\n")
cat("Churn rate in test :", percent(mean(test_df$churn  == "Yes"), 0.1), "\n")
```

```{r normalise}
# ── Fit the pre-processor on training data ONLY ───────────────────────────────
preproc <- preProcess(
  train_df %>% select(-churn),   # exclude the target
  method = c("center", "scale")  # center: subtract mean; scale: divide by sd
)

# ── Apply to both sets ────────────────────────────────────────────────────────
train_scaled <- predict(preproc, train_df)   # fit + transform on train
test_scaled  <- predict(preproc, test_df)    # transform only on test

# ── Verify: training feature means should be ~0, SDs ~1 ─────────────────────
post_stats <- train_scaled %>%
  select(-churn) %>%
  summarise(across(everything(), list(mean = mean, sd = sd)))

cat("Post-scaling training set verification:\n")
means <- train_scaled %>% select(-churn) %>% summarise(across(everything(), mean))
sds   <- train_scaled %>% select(-churn) %>% summarise(across(everything(), sd))
cat(sprintf("  Grand mean of all feature means: %.6f  (should be ~0)\n", mean(unlist(means))))
cat(sprintf("  Grand mean of all feature SDs  : %.6f  (should be ~1)\n", mean(unlist(sds))))
```

```{r normalisation-plot}
# Visualise before vs after normalisation for one feature
feat <- "longmon"

before_vals <- train_df[[feat]]
after_vals  <- train_scaled[[feat]]

par(mfrow = c(1, 2))

hist(before_vals,
     main   = paste(feat, "— Before Scaling"),
     xlab   = "Original value",
     col    = "steelblue",
     border = "white",
     breaks = 20)

hist(after_vals,
     main   = paste(feat, "— After StandardScaler"),
     xlab   = "Standardised z-score",
     col    = "tomato",
     border = "white",
     breaks = 20)

par(mfrow = c(1, 1))
```

---

# Step 6 — Train the Logistic Regression Model

**How logistic regression works in R:**

`glm(churn ~ ., family = binomial, data = train_scaled)`

- `family = binomial` → tells `glm` to use the **logit link function** (logistic regression)
- `churn ~ .` → predict `churn` using all other columns as predictors
- Under the hood, R uses **IRLS** (Iteratively Reweighted Least Squares) to fit the coefficients

```{r train-model}
# ── Train the logistic regression model ──────────────────────────────────────
# family = binomial(link = "logit") is the standard logistic regression specification
model <- glm(
  churn ~ .,
  data   = train_scaled,
  family = binomial(link = "logit")
)

cat("✅ Model trained successfully.\n\n")
cat("Model formula: logit(P(churn=Yes)) = β₀ + β₁·feature₁ + ... + βₙ·featureₙ\n")
```

```{r model-summary}
# Inspect the full model output
# ── Columns explained ────────────────────────────────────────────────────────
# Estimate  : β coefficient (log-odds unit; positive = increases churn odds)
# Std. Error: uncertainty in the estimate
# z value   : Estimate / Std.Error (Wald test statistic)
# Pr(>|z|)  : p-value — small → feature is statistically significant
summary(model)
```

```{r coefficient-plot}
# Visualise coefficients as a forest plot
coef_df <- broom::tidy(model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate)))

ggplot(coef_df, aes(x = estimate, y = reorder(term, abs(estimate)),
                     colour = estimate > 0)) +
  geom_vline(xintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.25) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
                      labels = c("Decreases churn", "Increases churn")) +
  labs(
    title    = "Logistic Regression Coefficients",
    subtitle = "Log-odds scale | Bars = 95% confidence interval",
    x = "Coefficient (log-odds)",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

---

# Step 7 — Model Evaluation

## 7a — Loss Function: Log-Loss (Binary Cross-Entropy)

The **loss function** is what the model minimises during training. For logistic regression, this is **log-loss** (also called binary cross-entropy):

$$\text{LogLoss} = -\frac{1}{n} \sum_{i=1}^{n} \left[ y_i \cdot \log(\hat{p}_i) + (1-y_i) \cdot \log(1-\hat{p}_i) \right]$$

- **Lower log-loss = better model**
- A random (uninformed) classifier gives log-loss ≈ **0.693** (= ln 2)
- A perfect model gives log-loss → **0**

We check it on both train and test to detect overfitting.

```{r log-loss}
# ── Get predicted probabilities ───────────────────────────────────────────────
# type = "response" returns P(churn=Yes) — probability on the [0,1] scale
# (using type="link" would return log-odds, not probabilities)

train_prob <- predict(model, newdata = train_scaled, type = "response")
test_prob  <- predict(model, newdata = test_scaled,  type = "response")

# Convert factor labels to 0/1 for metric functions
y_train_num <- as.numeric(train_scaled$churn) - 1
y_test_num  <- as.numeric(test_scaled$churn)  - 1

# ── Compute Log-Loss ──────────────────────────────────────────────────────────
# MLmetrics::LogLoss(predicted_probabilities, actual_labels)
train_logloss <- LogLoss(train_prob, y_train_num)
test_logloss  <- LogLoss(test_prob,  y_test_num)

cat("============================================\n")
cat(" LOSS FUNCTION — Binary Cross-Entropy (Log-Loss)\n")
cat("============================================\n")
cat(sprintf("  Training log-loss : %.4f\n", train_logloss))
cat(sprintf("  Test     log-loss : %.4f\n", test_logloss))
cat(sprintf("  Baseline (random) : 0.6931\n"))
cat("\n")
if (test_logloss < 0.5) {
  cat("  ✅ Model is substantially better than random.\n")
} else if (test_logloss < 0.65) {
  cat("  ⚠️  Model is modestly better than random.\n")
} else {
  cat("  ❌ Model near random — revisit feature selection.\n")
}
```

```{r roc-auc}
# ── AUC-ROC ──────────────────────────────────────────────────────────────────
# AUC = 1.0 → perfect ranking; AUC = 0.5 → random; AUC < 0.5 → worse than random

roc_obj <- roc(
  response  = test_scaled$churn,
  predictor = test_prob,
  levels    = c("No", "Yes"),
  direction = "<"    # higher probability → more likely "Yes" (churn)
)

auc_val <- auc(roc_obj)
cat(sprintf("AUC-ROC Score: %.4f\n", auc_val))

# Plot ROC curve
plot(
  roc_obj,
  col     = "tomato",
  lwd     = 2,
  main    = sprintf("ROC Curve — Churn Prediction  (AUC = %.3f)", auc_val),
  xlab    = "False Positive Rate (1 - Specificity)",
  ylab    = "True Positive Rate (Sensitivity / Recall)"
)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(sprintf("Logistic Regression (AUC=%.3f)", auc_val), "Random classifier"),
       col    = c("tomato", "grey50"),
       lwd    = c(2, 1),
       lty    = c(1, 2))
```

```{r predictions-and-confusion}
# ── Hard predictions using 0.5 threshold ─────────────────────────────────────
# If P(churn) > 0.5 → predict "Yes", else "No"
test_pred <- ifelse(test_prob > 0.5, "Yes", "No") %>%
  factor(levels = c("No", "Yes"))

# ── Confusion Matrix ──────────────────────────────────────────────────────────
# caret's confusionMatrix gives a rich summary with many derived metrics
cm <- confusionMatrix(
  data      = test_pred,
  reference = test_scaled$churn,
  positive  = "Yes"     # "Yes" = churn is the "positive" class
)

print(cm)
```

```{r confusion-plot}
# Visualise confusion matrix as a heatmap
cm_table <- as.data.frame(cm$table)
names(cm_table) <- c("Predicted", "Actual", "Count")

ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Count), size = 8, fontface = "bold", colour = "white") +
  scale_fill_gradient(low = "steelblue", high = "tomato") +
  labs(
    title    = "Confusion Matrix — Test Set",
    subtitle = "Rows = Predicted class | Columns = Actual class",
    fill     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

---

# Step 8 — Summary & Interpretation

```{r final-summary}
accuracy  <- cm$overall["Accuracy"]
precision <- cm$byClass["Precision"]
recall    <- cm$byClass["Recall"]
f1        <- cm$byClass["F1"]

cat("==============================================\n")
cat(" CHURN PREDICTION MODEL — FINAL SUMMARY\n")
cat("==============================================\n")
cat(sprintf("  Dataset           : %d customers, %d columns\n", nrow(df), ncol(df)))
cat(sprintf("  Features used     : %d\n", length(selected_features)))
cat(sprintf("  Train / Test split: 80%% / 20%% (stratified)\n\n"))
cat("  ─── Performance Metrics ───\n")
cat(sprintf("  Test Accuracy  : %.1f%%\n",  accuracy  * 100))
cat(sprintf("  Test Precision : %.1f%%\n",  precision * 100))
cat(sprintf("  Test Recall    : %.1f%%\n",  recall    * 100))
cat(sprintf("  Test F1-Score  : %.3f\n",    f1))
cat(sprintf("  Test Log-Loss  : %.4f  (lower = better)\n", test_logloss))
cat(sprintf("  Test AUC-ROC   : %.4f  (1.0 = perfect)\n", auc_val))
cat("\n")
cat("  ─── Steps Taken ───\n")
cat("  1. Loaded 200-customer telecom dataset\n")
cat("  2. Converted 'churn' to factor (categorical) for glm()\n")
cat("  3. Selected features with |Pearson r| > 0.10\n")
cat("  4. Removed collinear log-transformed duplicates\n")
cat("  5. Normalised with caret::preProcess (center + scale)\n")
cat("  6. Trained glm() with family=binomial(link='logit')\n")
cat("  7. Evaluated via log-loss, AUC-ROC, confusion matrix\n")
cat("==============================================\n")
```

---

# 📚 Key Concepts Recap

| Concept | R Tool Used | Why it matters |
|---|---|---|
| **Categorical response** | `factor(churn, levels=c(0,1))` | Required for `glm()` to run logistic regression |
| **StandardScaler** | `caret::preProcess("center","scale")` | Ensures IRLS optimiser converges properly |
| **Log-loss** | `MLmetrics::LogLoss()` | True training loss; penalises confident wrong predictions |
| **AUC-ROC** | `pROC::roc()` + `auc()` | Measures ranking quality regardless of threshold |
| **Confusion Matrix** | `caret::confusionMatrix()` | Shows TP/TN/FP/FN breakdown with many derived metrics |
| **Feature selection** | Pearson correlation, |r| > 0.10 | Reduces noise, improves model interpretability |

---

**Next steps to explore:**

- Use `class.weights` in `caret::train()` to handle class imbalance  
- Try **stepwise selection** with `step(model, direction="both")` for formal feature selection  
- Add L1 (LASSO) regularisation via `glmnet` package for automatic sparsity  
- Compare with `randomForest` or `xgboost` on the same feature set  

---

*End of notebook. Happy modelling! 🎉*
