---
title: "Customer Churn Prediction — Logistic Regression"
subtitle: "A Complete Teaching Notebook | From Raw Data to Confident Interpretation"
author: "Instructor Edition"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
# ─────────────────────────────────────────────────────────────────────────────
# GLOBAL CHUNK OPTIONS
#
# echo    = TRUE   — Always show code. This is a teaching notebook; every line
#                    of code is part of the lesson.
# message = FALSE  — Suppress package-load messages (informative but noisy).
# warning = FALSE  — Suppress warnings so the rendered HTML stays readable.
# fig.*            — Consistent figure dimensions across all plots.
# ─────────────────────────────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```

---

# Welcome — What This Notebook Teaches

This notebook teaches logistic regression from first principles using a
real-world telecom churn dataset. Every decision — why we convert the outcome
to a factor, why we normalise features, which loss function the model minimises
— is explained before any code is written. By the end you will know not just
*how* to run logistic regression in R, but *why* every step exists.

---

## The Business Problem

A telecom company needs to answer one question: which customers are likely to
cancel their subscription (churn)? Predicting churn before it happens allows
the business to intervene — offer discounts, improve service, or reach out
proactively — and protect revenue.

This is a **supervised binary classification** problem.

| Label | Meaning |
|-------|---------|
| `1` (becomes `"Yes"`) | Customer churned — cancelled their service |
| `0` (becomes `"No"`)  | Customer stayed — did not cancel |

---

## Why Logistic Regression, Not Linear Regression?

**Linear regression predicts a continuous number** — house price, temperature,
revenue. Forced to predict churn, it can output values such as `1.8` or `-0.3`,
which are not valid probabilities.

**Logistic regression** wraps the linear predictor inside a sigmoid function
that maps any real number into the interval (0, 1):

$$P(\text{churn} = 1 \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k)}}$$

Taking the log-odds (logit) of that probability recovers linearity:

$$\text{logit}(p) = \log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$

Logistic regression is a linear model — but it operates in log-odds space, and
the sigmoid back-transformation ensures every output is a valid probability.

---

## Roadmap

| Step | Task | Why It Matters |
|------|------|----------------|
| 1 | Load libraries | Make all tools available upfront |
| 2 | Load and explore data | Understand the data before touching the model |
| 3 | Cast response to factor | Required for `glm()` to apply logistic, not linear, regression |
| 4 | Select features thoughtfully | Remove noise; reduce multicollinearity |
| 5 | Normalise features | Guarantee optimiser convergence |
| 6 | Train / test split | Enable honest evaluation on unseen data |
| 7 | Train the model | Fit logistic regression with `glm()` |
| 8 | Evaluate with loss function and metrics | Measure model quality rigorously |
| 9 | Interpret coefficients | Understand what the model is actually saying |

---

# Step 1 — Load Libraries

Load all packages at the very top of the notebook. This makes every dependency
explicit and prevents cryptic "function not found" errors that appear mid-analysis
when a package was never loaded.

```{r load-libraries}
# ── Core data wrangling and visualisation ──────────────────────────────────────
library(tidyverse)     # dplyr, ggplot2, readr, tidyr, purrr, forcats
library(corrplot)      # heat-map style correlation matrix
library(scales)        # nice axis labels: percent(), comma(), dollar()

# ── Machine-learning pipeline ──────────────────────────────────────────────────
library(caret)         # createDataPartition(), preProcess(), confusionMatrix()

# ── Model evaluation ───────────────────────────────────────────────────────────
library(pROC)          # roc(), auc() — ROC curve and AUC-ROC metric
library(MLmetrics)     # LogLoss() — the loss function logistic regression minimises

# ── Coefficient interpretation ─────────────────────────────────────────────────
library(broom)         # tidy() — turns messy glm() output into a clean data frame

# ── Table formatting ───────────────────────────────────────────────────────────
library(knitr)         # kable() — renders clean HTML tables in the notebook
library(kableExtra)    # row/column-level styling for kable tables

cat("All libraries loaded successfully.\n")
```

---

# Step 2 — Load and Explore the Data

Never skip exploratory data analysis. A model is only as good as your
understanding of the data. EDA reveals missing values, unusual ranges, class
imbalance, and which features look promising — all before you touch the model.

## 2a — Load the Dataset

```{r load-data}
# read_csv() from readr (part of tidyverse) is faster and more type-consistent
# than base R's read.csv(). Critically, it never silently converts strings to
# factors, which avoids a whole class of subtle bugs.
#
# IMPORTANT: ChurnData.csv must be in your R working directory.
#   Check your current directory  :  getwd()
#   Change it                     :  setwd("path/to/your/folder")
#                                    or use the Files pane in RStudio.

df <- read_csv("ChurnData.csv")

cat("-----------------------------------\n")
cat(" DATASET OVERVIEW\n")
cat("-----------------------------------\n")
cat(sprintf("  Rows    : %d customers\n",  nrow(df)))
cat(sprintf("  Columns : %d features\n\n", ncol(df)))

# Display the first six rows as a formatted table for a quick sanity check
head(df) %>%
  kable(caption = "First 6 rows of ChurnData") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size  = 12
  )
```

## 2b — Data Types and Structure

```{r eda-structure}
# glimpse() is the tidyverse equivalent of str() or Python's df.info().
# It shows: column names, data types, and a short preview of values in each column.
#
# KEY OBSERVATION: All columns will likely show as <dbl> (numeric) —
# including 'churn'. That means 'churn' is stored as 0 / 1 integers,
# not as a category. We must fix this before modelling. Step 3 explains why.

glimpse(df)
```

## 2c — Summary Statistics

```{r eda-summary}
# summary() provides the five-number summary (min, Q1, median, Q3, max)
# plus the mean for every numeric column.
#
# Things to look for:
#   - Wide numeric ranges (e.g. 'income' running 0 to 1600 vs a binary 0/1 flag).
#     This confirms that normalisation is essential before model training (Step 5).
#   - Any column where Min == Max: zero variance, carries no useful information.
#   - NA counts appearing in any column: missing values must be handled first.

summary(df)
```

## 2d — Missing Value Audit

```{r eda-missing}
# Missing values corrupt model training silently if left unaddressed.
# We audit every column before proceeding.

missing_counts <- colSums(is.na(df))

if (sum(missing_counts) == 0) {
  cat("No missing values found. Clean dataset — we can proceed.\n")
} else {
  cat("Missing values detected:\n")
  missing_counts[missing_counts > 0] %>%
    as.data.frame() %>%
    setNames("Missing Count") %>%
    kable(caption = "Columns with missing values") %>%
    kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

---

# Step 3 — Cast the Response Variable to a Factor

**This is one of the most commonly missed steps. It changes the model fundamentally.**

## Why This Step Cannot Be Skipped

When `churn` is stored as numeric (0.0 / 1.0), R's `glm()` treats it as a
continuous outcome and fits **linear regression** — not logistic. The consequences
are severe:

- Predictions are not bounded to [0, 1]; values like -0.2 or 1.4 are possible.
- Coefficients are estimated by ordinary least squares, not maximum likelihood.
- The sigmoid link function is never applied.

Converting `churn` to a **factor** explicitly signals to `glm()`:
*"This is a categorical outcome with two levels — apply the binomial family
and the logit link function."*

```{r response-variable}
# ── BEFORE: numeric representation ────────────────────────────────────────────
cat("BEFORE casting:\n")
cat("  R class       :", class(df$churn), "\n")
cat("  Unique values :", paste(sort(unique(df$churn)), collapse = ", "), "\n\n")

# ── Cast to factor ─────────────────────────────────────────────────────────────
# levels = c(0, 1)         Explicit ordering — 0 becomes the reference level.
# labels = c("No", "Yes")  Human-readable labels (much clearer in model output).
#
# WHY put "No" first?
# R treats the FIRST factor level as the reference (baseline) category.
# Every coefficient therefore represents the change in log-odds of "Yes" (churn)
# relative to "No" (no churn). This is the standard, intuitive interpretation.

df$churn <- factor(df$churn, levels = c(0, 1), labels = c("No", "Yes"))

# ── AFTER: factor representation ──────────────────────────────────────────────
cat("AFTER casting:\n")
cat("  R class :", class(df$churn), "\n")
cat("  Levels  :", paste(levels(df$churn), collapse = " -> "), "\n\n")

cat("Class distribution:\n")
print(table(df$churn))
```

## Class Balance Check

```{r class-balance-plot}
# If 95% of customers are "No" and only 5% are "Yes", a naive model that
# always predicts "No" achieves 95% accuracy while being completely useless
# at identifying actual churners — the entire purpose of the model.
#
# A severe imbalance signals that:
#   - Accuracy is a misleading headline metric; prefer AUC-ROC and Log-Loss.
#   - Techniques like oversampling (SMOTE) or class weighting may be needed.

churn_counts       <- as.data.frame(table(df$churn))
names(churn_counts) <- c("Churn", "Count")
churn_counts$Pct   <- churn_counts$Count / sum(churn_counts$Count)
churn_counts$Label <- c("Did Not Churn (No)", "Churned (Yes)")

ggplot(churn_counts, aes(x = Label, y = Count, fill = Label)) +
  geom_col(width = 0.5, colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = paste0(Count, "\n(", percent(Pct, accuracy = 0.1), ")")),
    vjust = -0.4, fontface = "bold", size = 5
  ) +
  scale_fill_manual(
    values = c("Did Not Churn (No)" = "steelblue", "Churned (Yes)" = "tomato")
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Churn Class Distribution",
    subtitle = "Balanced is near 50/50. A highly skewed split requires careful metric choice.",
    x = NULL,
    y = "Number of Customers"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```

**Interpretation:** If the split is severely unequal, overall accuracy is a
misleading metric. This notebook evaluates the model with AUC-ROC and Log-Loss,
which are robust to class imbalance and tell a more honest story.

---

# Step 4 — Thoughtful Feature Selection

Including every column in your model is tempting but harmful. Irrelevant features
add noise. Correlated features distort coefficient estimates. Feature selection
is a deliberate craft, not an afterthought.

## 4a — Why Feature Selection Matters

| Problem | Consequence if ignored |
|---------|------------------------|
| Noise features | Model learns spurious patterns; generalises poorly to new data |
| Multicollinearity | Two correlated predictors share variance — coefficients become unstable and uninterpretable |
| Too many predictors | Risk of overfitting — model memorises training data, fails on test data |

## 4b — Point-Biserial Correlation With Churn

```{r feature-correlation}
# When one variable is binary (churn: 0/1) and the other is continuous,
# Pearson's r is mathematically equivalent to the point-biserial correlation
# — a valid measure of linear association between a continuous and a binary variable.
#
# Strategy: keep only features where |r| > 0.10 with churn.
# Features below this threshold have almost no linear relationship with the
# outcome and are unlikely to contribute meaningfully to the model.

# Temporarily restore churn to 0/1 numeric for the correlation calculation
churn_numeric <- as.numeric(df$churn) - 1   # "Yes"=2 -> 1, "No"=1 -> 0

# Identify all numeric predictor columns (churn itself excluded)
numeric_cols <- df %>%
  select(where(is.numeric)) %>%
  names()

# Compute the correlation of every numeric feature with churn
corr_with_churn <- sapply(numeric_cols, function(col) {
  cor(df[[col]], churn_numeric, use = "complete.obs")
})

# Build a sorted table — strongest absolute correlations at the top
corr_df <- tibble(
  Feature     = names(corr_with_churn),
  Correlation = corr_with_churn
) %>%
  arrange(desc(abs(Correlation)))

cat("Pearson correlation of each feature with 'churn' (sorted by absolute value):\n\n")

corr_df %>%
  mutate(
    Correlation = round(Correlation, 3),
    Strength = case_when(
      abs(Correlation) >= 0.30 ~ "Strong",
      abs(Correlation) >= 0.10 ~ "Moderate",
      TRUE                     ~ "Weak — will be dropped"
    )
  ) %>%
  kable(caption = "Feature-Churn Correlation (point-biserial)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(abs(corr_df$Correlation) <= 0.10), color = "grey")
```

## 4c — Visualise Correlations

```{r correlation-barplot}
# A horizontal bar chart makes the direction and magnitude of each relationship
# easy to read at a glance.

corr_df_plot <- corr_df %>%
  mutate(Direction = ifelse(Correlation > 0, "Positive", "Negative"))

ggplot(corr_df_plot,
       aes(x = Correlation,
           y = reorder(Feature, abs(Correlation)),
           fill = Direction)) +
  geom_col(colour = "white", linewidth = 0.4) +
  geom_vline(xintercept =  0.10, linetype = "dashed",
             colour = "forestgreen", linewidth = 0.9) +
  geom_vline(xintercept = -0.10, linetype = "dashed",
             colour = "forestgreen", linewidth = 0.9) +
  geom_vline(xintercept = 0, colour = "black", linewidth = 0.4) +
  scale_fill_manual(values = c("Negative" = "steelblue", "Positive" = "tomato")) +
  annotate("text", x =  0.12, y = 1.5, label = "Keep threshold (+0.10)",
           colour = "forestgreen", size = 3.2, hjust = 0) +
  annotate("text", x = -0.12, y = 1.5, label = "Keep threshold (-0.10)",
           colour = "forestgreen", size = 3.2, hjust = 1) +
  labs(
    title    = "Feature Correlation With Churn",
    subtitle = "Green dashed lines mark the selection threshold (|r| = 0.10). Features outside are retained.",
    x        = "Pearson Correlation Coefficient (r)",
    y        = NULL,
    fill     = "Direction"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 4d — Apply the Threshold

```{r select-features}
# Retain only features whose absolute correlation with churn exceeds the threshold.
# A constant of 0.10 is used here; in practice you may adjust this based on
# domain knowledge or cross-validation experiments.

THRESHOLD <- 0.10

selected_features <- corr_df %>%
  filter(abs(Correlation) > THRESHOLD) %>%
  pull(Feature)

cat(sprintf("Features retained (|r| > %.2f): %d out of %d total\n\n",
            THRESHOLD, length(selected_features), nrow(corr_df)))
cat("Selected features:\n")
cat(paste(" -", selected_features, collapse = "\n"), "\n")
```

## 4e — Multicollinearity Check Between Predictors

```{r multicollinearity-heatmap}
# So far we have examined how each feature correlates with CHURN.
# Now we check how the selected features correlate with EACH OTHER.
#
# Multicollinearity occurs when two predictor variables are highly correlated
# with one another. Including both:
#   - Makes coefficient estimates unstable (high variance)
#   - Inflates standard errors, making p-values unreliable
#   - Prevents the model from distinguishing which variable is doing the work
#
# Rule of thumb: if |r| > 0.80 between two predictors, drop one of them.

feat_corr_matrix <- df %>%
  select(all_of(selected_features)) %>%
  cor(use = "complete.obs")

corrplot(
  feat_corr_matrix,
  method      = "color",
  type        = "lower",
  tl.cex      = 0.75,
  tl.col      = "black",
  addCoef.col = "black",
  number.cex  = 0.55,
  col         = colorRampPalette(c("steelblue", "white", "tomato"))(200),
  title       = "Feature-Feature Correlation Matrix (Multicollinearity Check)",
  mar         = c(0, 0, 2, 0)
)
```

## 4f — Remove Collinear Duplicates

```{r remove-collinear}
# Reading the heatmap above:
#
# 'loglong' is the natural-log transform of 'longmon'  — same information, different scale
# 'lninc'   is the natural-log transform of 'income'   — same information, different scale
#
# Including both versions of the same information inflates coefficient variance
# and makes the model harder to interpret.
#
# DECISION: Keep the original untransformed variables (longmon, income) because:
#   - They are on interpretable scales (minutes of calls; dollars of income).
#   - After normalisation in Step 5, the log transforms offer no additional benefit.

features_to_drop  <- c("loglong", "lninc")
selected_features <- setdiff(selected_features, features_to_drop)

cat(sprintf("Final feature set: %d features (removed %d log-transform duplicates)\n\n",
            length(selected_features), length(features_to_drop)))

corr_df %>%
  filter(Feature %in% selected_features) %>%
  arrange(desc(abs(Correlation))) %>%
  mutate(
    Correlation    = round(Correlation, 3),
    Interpretation = ifelse(
      Correlation > 0,
      "Higher value -> more likely to churn",
      "Higher value -> less likely to churn"
    )
  ) %>%
  kable(caption = "Final selected features with churn correlation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Step 5 — Normalisation (Ensuring Convergence)

Normalisation is one of the most frequently skipped steps by beginners — and one
of the most consequential.

## Why Normalisation Is Essential for Logistic Regression

R's `glm()` fits logistic regression using **Iteratively Reweighted Least Squares
(IRLS)**, an iterative algorithm that takes gradient steps — small coefficient
updates at each iteration — guided by the curvature of the log-likelihood surface.

**The problem:** If `income` ranges from 0 to 1,600 and a binary flag sits at 0
or 1, the gradient updates for each feature operate on completely different scales.
The optimiser either takes infinitesimally small steps for large-scale features
(very slow convergence) or overshoots for small-scale features (oscillation or
outright failure to converge).

**The solution — z-score standardisation:**

$$z = \frac{x - \mu_{\text{train}}}{\sigma_{\text{train}}}$$

After scaling every feature has mean = 0 and standard deviation = 1. The optimiser
works in a uniform space and converges reliably.

**Critical rule — no data leakage:**  
The scaler must be *fitted on training data only*, then *applied* to both sets.
If you fit on the full dataset (including the test set), test-set information
bleeds into the training process — a mistake called data leakage that
artificially inflates evaluation metrics.

## 5a — Train / Test Split First

```{r train-test-split}
# ── Prepare the modelling dataset ─────────────────────────────────────────────
# Select only the retained features plus the target variable.

model_df <- df %>%
  select(all_of(c(selected_features, "churn")))

# ── Stratified 80 / 20 split ──────────────────────────────────────────────────
# "Stratified" means the churn proportion is preserved in BOTH the training
# and test sets. Without stratification, the test set might — by random chance
# — contain very few churners, making evaluation unreliable.
#
# set.seed(42) ensures reproducibility: every run of this notebook produces
# exactly the same training and test rows.

set.seed(42)

train_idx <- createDataPartition(
  model_df$churn,
  p     = 0.80,    # 80% goes to training
  list  = FALSE,
  times = 1
)

train_df <- model_df[ train_idx, ]   # rows IN the partition -> training set
test_df  <- model_df[-train_idx, ]   # rows NOT in partition -> test set

cat("-----------------------------------\n")
cat(" TRAIN / TEST SPLIT\n")
cat("-----------------------------------\n")
cat(sprintf("  Training samples : %d (%.0f%%)\n",
            nrow(train_df), 100 * nrow(train_df) / nrow(model_df)))
cat(sprintf("  Test samples     : %d (%.0f%%)\n",
            nrow(test_df),  100 * nrow(test_df)  / nrow(model_df)))
cat(sprintf("\n  Churn rate — train : %s\n",
            percent(mean(train_df$churn == "Yes"), accuracy = 0.1)))
cat(sprintf("  Churn rate — test  : %s  (should be close to train if stratified)\n",
            percent(mean(test_df$churn  == "Yes"), accuracy = 0.1)))
```

## 5b — Fit the Scaler and Apply

```{r normalise}
# ── Step 1: Fit preProcess parameters on the TRAINING SET ONLY ────────────────
# "center" = subtract the training mean from each feature column.
# "scale"  = divide by the training standard deviation.
# Together these implement z-score standardisation.
#
# The target variable (churn) must be excluded — it must not be transformed.

preproc <- preProcess(
  train_df %>% select(-churn),
  method = c("center", "scale")
)

# ── Step 2: Apply the SAME parameters to both sets ────────────────────────────
# predict() here does not make predictions — it applies the fitted transformation.
# This naming follows the standard caret interface for preProcess objects.

train_scaled <- predict(preproc, train_df)
test_scaled  <- predict(preproc, test_df)

# ── Step 3: Verify the scaling worked ─────────────────────────────────────────
means_after <- train_scaled %>% select(-churn) %>% summarise(across(everything(), mean))
sds_after   <- train_scaled %>% select(-churn) %>% summarise(across(everything(), sd))

cat("Post-scaling verification (training set):\n")
cat(sprintf("  Grand mean of all feature means : %.8f  (target: 0)\n",
            mean(unlist(means_after))))
cat(sprintf("  Grand mean of all feature SDs   : %.8f  (target: 1)\n",
            mean(unlist(sds_after))))
cat("\nScaling confirmed. The optimiser will now work in a uniform space.\n")
```

## 5c — Visualise the Effect of Scaling

```{r normalisation-plot}
# Side-by-side histograms for 'longmon' before and after scaling.
#
# The key teaching point: the SHAPE of the distribution is identical in both
# panels. Only the axis units change — from raw minutes to standard deviations
# from the mean. Scaling re-expresses data in a common unit; it does not distort
# relationships or alter the relative ordering of observations.

feat <- "longmon"

before_vals <- train_df[[feat]]
after_vals  <- train_scaled[[feat]]

par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

hist(before_vals,
     main   = paste(feat, "- BEFORE Scaling"),
     xlab   = sprintf("Original value  (mean = %.1f, SD = %.1f)",
                      mean(before_vals), sd(before_vals)),
     col    = "steelblue",
     border = "white",
     breaks = 20)
abline(v = mean(before_vals), col = "red", lwd = 2, lty = 2)

hist(after_vals,
     main   = paste(feat, "- AFTER Scaling"),
     xlab   = sprintf("Z-score  (mean = %.4f, SD = %.4f)",
                      mean(after_vals), sd(after_vals)),
     col    = "tomato",
     border = "white",
     breaks = 20)
abline(v = 0, col = "darkred", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

---

# Step 6 — Train the Logistic Regression Model

Once the data is correctly prepared, the modelling step itself is concise.
The hard work was in Steps 3 through 5.

## How glm() Fits Logistic Regression

| Argument | Meaning |
|----------|---------|
| `churn ~ .` | Predict `churn` using all remaining columns as predictors |
| `family = binomial` | The outcome is binary — use the binomial family |
| `link = "logit"` | Apply the log-odds link function — the standard for binary outcomes |

Internally, R uses IRLS: it repeatedly adjusts coefficients to maximise the
log-likelihood of observing the training labels given the current parameter
estimates. Convergence is reached when the change in log-likelihood between
iterations falls below a small tolerance.

```{r train-model}
# ── Fit the logistic regression model ─────────────────────────────────────────
model <- glm(
  churn ~ .,
  data   = train_scaled,
  family = binomial(link = "logit")
)

cat("Model trained successfully.\n\n")
cat("What this model learns:\n")
cat("  logit(P(churn = Yes)) = b0 + b1*x1 + b2*x2 + ... + bk*xk\n\n")
cat("  where logit(p) = log(p / (1-p))  -- the log of the odds ratio\n")
cat("  To recover a probability: P = 1 / (1 + exp(-linear predictor))\n")
```

## 6b — Raw Model Summary

```{r model-summary}
# Reading the glm() summary output:
#
#   Estimate   : beta coefficient in LOG-ODDS units.
#                Positive  => feature increases the odds of churn.
#                Negative  => feature decreases the odds of churn.
#
#   Std. Error : uncertainty around the estimate.
#                A large standard error signals low confidence in that coefficient.
#
#   z value    : Estimate / Std. Error (Wald test statistic).
#                A large absolute z value means the coefficient is far from zero.
#
#   Pr(>|z|)   : p-value for the Wald test.
#                < 0.05  (*)   significant at the 5% level
#                < 0.01  (**)  significant at the 1% level
#                < 0.001 (***) significant at the 0.1% level

summary(model)
```

## 6c — Odds Ratio Interpretation

```{r odds-ratios}
# Log-odds coefficients are mathematically convenient but hard to interpret
# intuitively. Exponentiating them gives ODDS RATIOS, which are much easier
# to communicate:
#
#   OR > 1  =>  feature INCREASES churn odds
#   OR < 1  =>  feature DECREASES churn odds
#   OR = 1  =>  feature has no effect on churn odds
#
# Example: OR = 1.50 for 'longmon' means a 1-SD increase in long-distance
# minutes is associated with 50% higher odds of churning, all else held constant.

coef_table <- tidy(model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OddsRatio  = round(exp(estimate), 3),
    OR_CI_Low  = round(exp(conf.low),  3),
    OR_CI_High = round(exp(conf.high), 3),
    p.value    = round(p.value, 4),
    Significant = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "** ",
      p.value < 0.05  ~ "*  ",
      TRUE            ~ "   "
    ),
    Effect = ifelse(OddsRatio > 1, "Increases churn odds", "Decreases churn odds")
  ) %>%
  select(term, OddsRatio, OR_CI_Low, OR_CI_High, p.value, Significant, Effect) %>%
  arrange(desc(OddsRatio))

coef_table %>%
  kable(caption = "Odds Ratios — Effect of each feature on churn odds (per 1 SD change)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(7, bold = TRUE,
              color = ifelse(coef_table$Effect == "Increases churn odds",
                             "tomato", "steelblue"))
```

## 6d — Coefficient Forest Plot

```{r coefficient-plot}
# A forest plot shows each coefficient in log-odds space with its 95%
# confidence interval bar.
#
# If a confidence interval bar CROSSES ZERO, the effect may not be
# statistically reliable. Zero in log-odds corresponds to an odds ratio
# of 1, meaning no effect on churn odds.

coef_df <- tidy(model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate)))

ggplot(coef_df, aes(x = estimate,
                     y = reorder(term, abs(estimate)),
                     colour = estimate > 0)) +
  geom_vline(xintercept = 0, colour = "grey50", linetype = "dashed", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_colour_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
    labels = c("FALSE" = "Decreases churn odds", "TRUE" = "Increases churn odds")
  ) +
  labs(
    title    = "Logistic Regression Coefficients (Log-Odds Scale)",
    subtitle = "Error bars show 95% confidence intervals. Bars crossing zero are statistically uncertain.",
    x        = "Coefficient (log-odds)",
    y        = NULL,
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

---

# Step 7 — Model Evaluation

Accuracy alone is a poor metric for imbalanced classification. We use several
complementary measures to build a complete picture of model quality.

## 7a — The Loss Function: Log-Loss (Binary Cross-Entropy)

The **loss function** is the quantity logistic regression *minimises* during
training. For binary classification this is **log-loss**, also called binary
cross-entropy:

$$\mathcal{L} = -\frac{1}{n} \sum_{i=1}^{n} \Bigl[ y_i \cdot \log(\hat{p}_i) + (1 - y_i) \cdot \log(1 - \hat{p}_i) \Bigr]$$

**Intuition:** log-loss heavily penalises *confident wrong predictions*. If the
model assigns P(churn) = 0.99 to a customer who did not churn, that single
observation contributes a large penalty. The model therefore learns to be
well-calibrated — not merely directionally correct, but appropriately uncertain
when the signal is weak.

| Log-Loss value | Interpretation |
|----------------|----------------|
| Close to 0 | Nearly perfect model |
| 0.693 (ln 2) | Random classifier — coin-flip baseline |
| Greater than 0.693 | Worse than random guessing |

```{r log-loss}
# ── Obtain predicted probabilities ────────────────────────────────────────────
# type = "response" returns P(churn = "Yes") in [0, 1].
# type = "link" would return raw log-odds — not what metric functions expect.

train_prob <- predict(model, newdata = train_scaled, type = "response")
test_prob  <- predict(model, newdata = test_scaled,  type = "response")

# Convert factor labels to 0 / 1 numeric for the metric functions
y_train_num <- as.numeric(train_scaled$churn) - 1   # "Yes" -> 1, "No" -> 0
y_test_num  <- as.numeric(test_scaled$churn)  - 1

# ── Compute Log-Loss ──────────────────────────────────────────────────────────
# MLmetrics::LogLoss(y_pred, y_true)  -- predictions first, true labels second
train_logloss <- LogLoss(train_prob, y_train_num)
test_logloss  <- LogLoss(test_prob,  y_test_num)

cat("-------------------------------------------\n")
cat(" LOSS FUNCTION — Binary Cross-Entropy\n")
cat("-------------------------------------------\n")
cat(sprintf("  Training log-loss       : %.4f\n", train_logloss))
cat(sprintf("  Test log-loss           : %.4f\n", test_logloss))
cat(sprintf("  Random baseline (ln 2)  : 0.6931\n"))
cat(sprintf("  Improvement vs baseline : %.1f%%\n\n",
            100 * (0.6931 - test_logloss) / 0.6931))

# Overfitting diagnostic
gap <- test_logloss - train_logloss
cat(sprintf("  Train-test gap          : %.4f", gap))
if (gap > 0.05) {
  cat("  — possible overfitting: test loss is notably higher than train\n")
} else {
  cat("  — model generalises well: train and test loss are close\n")
}

cat("\n")
if      (test_logloss < 0.40) {
  cat("  Rating: Excellent — substantially better than random\n")
} else if (test_logloss < 0.55) {
  cat("  Rating: Good — meaningfully better than random\n")
} else if (test_logloss < 0.65) {
  cat("  Rating: Modest — only slightly better than random\n")
} else {
  cat("  Rating: Poor — near random; revisit feature selection\n")
}
```

## 7b — AUC-ROC: Ranking Quality

```{r roc-auc}
# The ROC (Receiver Operating Characteristic) curve plots the True Positive Rate
# (Sensitivity) against the False Positive Rate at every possible probability
# threshold from 0 to 1.
#
# AUC = Area Under the ROC Curve
#   AUC = 1.0  =>  perfect — every churner is ranked above every non-churner
#   AUC = 0.5  =>  random — no discrimination ability at all
#   AUC < 0.5  =>  worse than random (the model's ranking is effectively inverted)
#
# Key property: AUC is threshold-independent. It measures how well the model
# RANKS customers by churn risk, regardless of where you set the decision cutoff.
# This makes AUC robust and meaningful on imbalanced datasets.

roc_obj <- roc(
  response  = test_scaled$churn,
  predictor = test_prob,
  levels    = c("No", "Yes"),
  direction = "<"   # higher probability corresponds to the positive class ("Yes")
)

auc_val <- auc(roc_obj)

cat(sprintf("AUC-ROC: %.4f\n\n", auc_val))
cat(sprintf("Interpretation: the model correctly ranks a randomly chosen churner\n"))
cat(sprintf("above a randomly chosen non-churner %.1f%% of the time.\n", auc_val * 100))

# ── Plot the ROC curve ─────────────────────────────────────────────────────────
plot(
  roc_obj,
  col  = "tomato",
  lwd  = 2.5,
  main = sprintf("ROC Curve — Churn Prediction  |  AUC = %.4f", auc_val),
  xlab = "False Positive Rate (1 - Specificity)",
  ylab = "True Positive Rate (Sensitivity / Recall)"
)
# Diagonal line = random classifier baseline (AUC = 0.5)
abline(a = 0, b = 1, lty = 2, col = "grey60", lwd = 1.5)

legend(
  "bottomright",
  legend = c(sprintf("Logistic Regression (AUC = %.3f)", auc_val),
             "Random Classifier  (AUC = 0.500)"),
  col    = c("tomato", "grey60"),
  lwd    = c(2.5, 1.5),
  lty    = c(1, 2),
  bty    = "n"
)

# Shade the area under the curve for visual emphasis
polygon(
  c(roc_obj$specificities, 0),
  c(roc_obj$sensitivities, 0),
  col    = adjustcolor("tomato", alpha.f = 0.12),
  border = NA
)
```

## 7c — Confusion Matrix and Classification Metrics

```{r predictions-and-confusion}
# ── Convert probabilities to class labels ─────────────────────────────────────
# At threshold = 0.5: if P(churn) > 0.5 predict "Yes", otherwise predict "No".
#
# 0.5 is the default threshold, not necessarily the optimal one. In practice
# you might lower it (e.g. to 0.3) to catch more actual churners at the cost
# of more false alarms. This is a business decision, not a model decision.

test_pred <- ifelse(test_prob > 0.5, "Yes", "No") %>%
  factor(levels = c("No", "Yes"))

# ── Confusion Matrix ───────────────────────────────────────────────────────────
# The four cells:
#
#   True Positive  (TP): predicted "Yes", actual "Yes"  — correctly caught churner
#   True Negative  (TN): predicted "No",  actual "No"   — correctly retained customer
#   False Positive (FP): predicted "Yes", actual "No"   — unnecessary intervention
#   False Negative (FN): predicted "No",  actual "Yes"  — missed churner (often costly)
#
# positive = "Yes" tells caret which class is the event of interest, so that
# Sensitivity, Precision, and Recall are all computed with respect to churn.

cm <- confusionMatrix(
  data      = test_pred,
  reference = test_scaled$churn,
  positive  = "Yes"
)

print(cm)
```

## 7d — Confusion Matrix Heatmap

```{r confusion-plot}
# A heatmap version of the confusion matrix is easier to scan at a glance
# than a raw numeric table.

cm_table       <- as.data.frame(cm$table)
names(cm_table) <- c("Predicted", "Actual", "Count")

total          <- sum(cm_table$Count)
cm_table$Pct   <- percent(cm_table$Count / total, accuracy = 0.1)
cm_table$Label <- paste0(cm_table$Count, "\n(", cm_table$Pct, ")")

ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = Label), size = 7, fontface = "bold", colour = "white") +
  scale_fill_gradient(low = "steelblue", high = "tomato") +
  labs(
    title    = "Confusion Matrix — Test Set",
    subtitle = "Rows = Predicted class | Columns = Actual class",
    fill     = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

## 7e — Complete Metrics Summary

```{r metrics-summary-table}
# Collect all key metrics from the confusionMatrix object into one summary table.
# Each metric answers a different question about model quality.

accuracy    <- cm$overall["Accuracy"]
precision   <- cm$byClass["Precision"]    # of all predicted churners, how many actually churned?
recall      <- cm$byClass["Recall"]       # of all actual churners, how many did we catch?
f1          <- cm$byClass["F1"]           # harmonic mean of precision and recall
specificity <- cm$byClass["Specificity"]  # of all actual non-churners, how many correctly labelled?

metrics_table <- tibble(
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall (Sensitivity)",
    "Specificity",
    "F1-Score",
    "Log-Loss",
    "AUC-ROC"
  ),
  Value = c(
    percent(accuracy,    accuracy = 0.1),
    percent(precision,   accuracy = 0.1),
    percent(recall,      accuracy = 0.1),
    percent(specificity, accuracy = 0.1),
    round(f1,           4),
    round(test_logloss, 4),
    round(auc_val,      4)
  ),
  Interpretation = c(
    "Percentage of all predictions that are correct",
    "Of all predicted churners, how many actually churned?",
    "Of all actual churners, what fraction did we catch?",
    "Of all actual non-churners, what fraction did we correctly identify?",
    "Harmonic mean of Precision and Recall — 1.0 is perfect",
    "Lower is better | 0.6931 = coin-flip baseline",
    "Higher is better | 0.5 = coin-flip | 1.0 = perfect"
  )
)

metrics_table %>%
  kable(caption = "Complete Model Evaluation — Test Set") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(5:7, bold = TRUE, background = "#f0f4f8")
```

**Which metric matters most in a churn problem?**  
Recall (catching actual churners) is typically more important than Precision.
A missed churner represents permanently lost revenue; a false alarm costs one
retention call. The business context determines which metric to prioritise,
and the optimal threshold can be shifted on the ROC curve to reflect that priority.

---

# Step 8 — Final Summary

```{r final-summary}
cat("======================================================\n")
cat("  CHURN PREDICTION MODEL — COMPLETE SUMMARY\n")
cat("======================================================\n\n")

cat("  DATA\n")
cat(sprintf("  |- Dataset          : %d customers, %d total columns\n", nrow(df), ncol(df)))
cat(sprintf("  |- Features used    : %d (after correlation filtering)\n", length(selected_features)))
cat(sprintf("  '- Train/Test split : 80%%/20%% stratified (seed = 42)\n\n"))

cat("  PREPROCESSING\n")
cat("  |- Response variable cast to factor (required for glm binomial)\n")
cat("  |- Features filtered: |Pearson r| > 0.10 with churn\n")
cat("  |- Log-transform duplicates removed: loglong, lninc\n")
cat("  '- StandardScaler applied (center + scale, fitted on train only)\n\n")

cat("  MODEL\n")
cat("  '- glm(family = binomial(link = 'logit')) — Logistic Regression\n\n")

cat("  PERFORMANCE — TEST SET\n")
cat(sprintf("  |- Accuracy    : %s\n",  percent(accuracy,    accuracy = 0.1)))
cat(sprintf("  |- Precision   : %s\n",  percent(precision,   accuracy = 0.1)))
cat(sprintf("  |- Recall      : %s\n",  percent(recall,      accuracy = 0.1)))
cat(sprintf("  |- Specificity : %s\n",  percent(specificity, accuracy = 0.1)))
cat(sprintf("  |- F1-Score    : %.4f\n",  f1))
cat(sprintf("  |- Log-Loss    : %.4f  (random baseline: 0.6931)\n", test_logloss))
cat(sprintf("  '- AUC-ROC     : %.4f  (random baseline: 0.5000)\n", auc_val))
cat("======================================================\n")
```

---

# Concepts Recap

| Concept | R Tool | Why It Matters |
|---------|--------|----------------|
| Binary categorical outcome | `factor(churn, levels = c(0,1))` | Required for `glm()` to apply logistic, not linear, regression |
| Sigmoid function | Built into `glm(family = binomial)` | Maps the linear predictor to a valid probability in (0, 1) |
| Feature selection | Pearson correlation, threshold `|r| > 0.10` | Removes noise and controls multicollinearity |
| Multicollinearity check | `corrplot()` heatmap | Identifies redundant predictors to remove |
| z-score normalisation | `caret::preProcess("center", "scale")` | Ensures the IRLS optimiser converges reliably |
| No data leakage | Scaler fitted on training set only | Prevents inflated test-set performance metrics |
| Log-Loss | `MLmetrics::LogLoss()` | The actual loss function — penalises confident wrong predictions |
| AUC-ROC | `pROC::roc()` and `auc()` | Threshold-independent ranking quality — robust to class imbalance |
| Confusion Matrix | `caret::confusionMatrix()` | Exposes TP / TN / FP / FN and all derived classification metrics |
| Odds Ratios | `exp(coef(model))` | Converts log-odds coefficients into interpretable effect sizes |

---

# Suggested Next Steps

**1. Handle class imbalance with class weights**

```r
# Penalise missing a churner more heavily than a false alarm.
model_weighted <- glm(
  churn ~ .,
  data    = train_scaled,
  family  = binomial,
  weights = ifelse(train_scaled$churn == "Yes", 3, 1)
)
```

**2. Optimise the classification threshold**

```r
# The default 0.5 threshold is rarely optimal.
# Sweep across candidate thresholds and select the one that maximises
# the metric most relevant to the business (e.g. Recall or F1-Score).
thresholds <- seq(0.1, 0.9, by = 0.05)
```

**3. Regularised logistic regression (LASSO / Ridge)**

```r
library(glmnet)
# L1 penalty (LASSO) automatically shrinks unimportant coefficients to zero,
# providing built-in feature selection and reducing the risk of overfitting.
```

**4. Stepwise feature selection guided by AIC**

```r
# AIC balances model fit against complexity.
# step() adds or removes predictors greedily to minimise AIC.
model_step <- step(model, direction = "both")
```

**5. Compare against non-linear models**

```r
library(randomForest)
library(xgboost)
# Tree-based ensemble methods capture non-linear relationships and feature
# interactions that logistic regression cannot express. Comparing results
# tells you whether the linear assumptions are adequate for this dataset.
```

---

*End of notebook. You now have a rigorous, fully documented foundation in logistic
regression — covering data preparation, feature selection, normalisation, model
training, loss-function evaluation, and coefficient interpretation. Every concept
here transfers directly to any binary classification problem.*
